---
title: "MRKT_UberCase"
author: "Jingyuan Wang"
date: "05/02/2022"
output: word_document
---
```{r}
library(dygraphs)
library(xts)
library(dplyr)
library(ggplot2)
library(lubridate)
library(rpart)
```

## Data Acquisation
```{r}
setwd("~/Documents/McGill/MRKT671/A2")
uber_data = read.csv('uber_data.csv')
head(uber_data)
```
```{r}
summary(uber_data)
```
Convert start time to datetime object
```{r}
uber_data[['period_start']] <- as.POSIXct(uber_data[['period_start']],
                                   format = "%Y-%m-%d %H:%M")
```

Drop columns that contains repetitive values
```{r}
uber_data = select(uber_data, -c(wait_time,city_id))
head(uber_data)
```

Set categorical variables as factors
```{r}
factor_cols = c('treat','commute')
uber_data[factor_cols] <- lapply(uber_data[factor_cols], factor)
```

```{r}
summary(uber_data)
```

### Define Target Effects
What is the effect of extending wait times from two to five minutes on:
2 vs 5 (primary)
products (good to have)
- the total number of shared rides completed (that is, rides taken via both the existing shared rides product—UberPOOL and the new shared rides product, Express), 
- the proportion of shared rides that were matched, 
- and driver payout per trip

# Weather consideration

```{r}
uber_data$total_trips = uber_data$trips_pool + uber_data$trips_express
uber_data$matched_prop = uber_data$total_matches/uber_data$total_trips
uber_data$avg_ride_payout = uber_data$total_driver_payout/uber_data$total_trips
head(uber_data)
```
## Data Discovery
### Trips
Total trips
```{r}
ggplot(uber_data,aes(period_start,total_trips))+geom_line(aes(color="Total Trips"))+labs(color="Data Type")
#plot(x = 'period_start', y = 'total_trips', data = uber_data, main = "Total Trips")
```


```{r}
treatment_uber = uber_data[uber_data$treat == TRUE,] #5 min
control_uber = uber_data[uber_data$treat == FALSE,] #2 min
```

Note that the experiment is performed on a alternative fashion.
```{r}
merge_data = merge(x = treatment_uber, y = control_uber, by = 'period_start', all = TRUE)
head(merge_data)
```
This means we need to encode the time series information in order to capture the periodicity underlying the data set.

Compare between # pool rides
```{r}
ggplot(treatment_uber,aes(period_start,trips_pool))+geom_line(aes(color="trips_pool_5min"))+geom_line(data=control_uber,aes(color="trips_pool_2min"))+labs(color="service type")+geom_point(data=treatment_uber[treatment_uber$commute==TRUE,],aes(period_start,trips_pool))+ geom_point(data=control_uber[control_uber$commute==TRUE,],aes(period_start,trips_pool))
```

In February, we noticed that patterns in treatment and condition group are relatively similar. However as time moved towards beginning of March, number of pool rides in treatment group surpassed condition group.

Note that weather might play a very important role here.


```{r}
ggplot(treatment_uber,aes(period_start,trips_express))+geom_line(aes(color="trips_express_5min"))+geom_line(data=control_uber,aes(color="trips_express_2min"))+labs(color="service type")+geom_point(data=treatment_uber[treatment_uber$commute==TRUE,],aes(period_start,trips_express))+ geom_point(data=control_uber[control_uber$commute==TRUE,],aes(period_start,trips_express))
```

With Express service, both groups present similar patterns.

### Pool and express service gaps
Control: 2 min
```{r}
don_control <- xts(x = control_uber[,c('trips_pool', 'trips_express')], order.by = control_uber$period_start)
p_control <- dygraph(don_control, main = "2 Minutes Wait Time Service Gap")
p_control
```

For condition group, the gap between two services are consistent.

Treatment: 5 min
```{r}
don_trt <- xts(x = treatment_uber[,c('trips_pool', 'trips_express')], order.by = treatment_uber$period_start)
p_trt <- dygraph(don_trt, main="5 Minutes Wait Time Service Gap")
p_trt
```
For treatment group, gap between two services seems to reduce.

## Deltas between pool and express
```{r}
colnames(uber_data)
```

```{r}
uber_data$service_deltas = uber_data$trips_express - uber_data$trips_pool
```

5 min
```{r}
trt_deltas = uber_data[uber_data$treat == TRUE,]
cond_deltas = uber_data[uber_data$treat == FALSE,]
ggplot(trt_deltas,aes(period_start,service_deltas))+geom_line(aes(color="service_delta_5min"))+geom_smooth(method="auto", se=TRUE, fullrange=FALSE, level=0.95)+labs(color="service type")
```
2 min
```{r}
ggplot(cond_deltas,aes(period_start,service_deltas))+geom_line(aes(color="service_delta_2min"))+geom_smooth(method="auto", se=TRUE, fullrange=FALSE, level=0.95)+labs(color="service type")
```
Both groups has a decreasing trend in service gaps as time moved into March. However, longer wait time tend to have a smaller gap.

### Cancellation
```{r}
ggplot(treatment_uber,aes(period_start,rider_cancellations))+geom_line(aes(color="rider_cancellations_5min"))+geom_line(data=control_uber,aes(color="rider_cancellations_2min"))+labs(color="service type")+geom_point(data=treatment_uber[treatment_uber$commute==TRUE,],aes(period_start,rider_cancellations))+ geom_point(data=control_uber[control_uber$commute==TRUE,],aes(period_start,rider_cancellations))
```

### Matches (>1 customers vs >2 customers)
Treatment (5 min)
```{r}
don_matches <- xts(x = treatment_uber[,c('total_matches', 'total_double_matches')], order.by = treatment_uber$period_start)
p <- dygraph(don_matches) 
p
```
Condition (2 min)
```{r}
don_matches_control <- xts(x = control_uber[,c('total_matches', 'total_double_matches')], order.by = control_uber$period_start)
p <- dygraph(don_matches_control)
p
```

Differences between matches are consistent in both groups.

### Proportion on matched trips
```{r}
colnames(treatment_uber)
```


```{r}
ggplot(treatment_uber,aes(period_start,matched_prop))+geom_line(aes(color="matched_prop_treatment"))+geom_line(data=control_uber,aes(color="matched_prop_condition"))+labs(color="service type")+geom_point(data=treatment_uber[treatment_uber$commute==TRUE,],aes(period_start,matched_prop))+ geom_point(data=control_uber[control_uber$commute==TRUE,],aes(period_start,matched_prop))
```

Rides happened during rush hours tend to have a higher matched proportion.

On average, treatment group (5 min wait time) has a lower proportion of matching.

### Total trips treatment vs condition
```{r}
ggplot(treatment_uber,aes(period_start,total_trips))+geom_line(aes(color="Total_Trips_5min"))+geom_line(data=control_uber,aes(color="Total_Trips_2min"))+labs(color="service type")+geom_point(data=treatment_uber[treatment_uber$commute==TRUE,],aes(period_start,total_trips))+ geom_point(data=control_uber[control_uber$commute==TRUE,],aes(period_start,total_trips))
```

### Driver's payout per trip
```{r}
ggplot(treatment_uber,aes(period_start,avg_ride_payout))+geom_line(aes(color="payout_treatment"))+geom_line(data=control_uber,aes(color="payout_condition"))+labs(color="service type")+geom_point(data=treatment_uber[treatment_uber$commute==TRUE,],aes(period_start,avg_ride_payout))+ geom_point(data=control_uber[control_uber$commute==TRUE,],aes(period_start,avg_ride_payout))
```

As time proceeds, we notice payout for condition group (2 min wait time) tends to become higher than treatment group.

## Seasonality and time series encoding
Traffic data tends to have daily and weekly periodicity. To deal with rotational experiment, we will encode the time-series datetime into three variables, month, weekday and hour.
```{r}
uber_data$period_month = month(as.POSIXlt(uber_data$period_start))
uber_data$period_weekday = as.POSIXlt(uber_data$period_start)$wday
uber_data$period_hour = as.POSIXlt(uber_data$period_start)$hour
```

```{r}
date_factor_cols = c('period_month','period_weekday', 'period_hour')
uber_data[date_factor_cols] <- lapply(uber_data[date_factor_cols], factor)
```

```{r}
summary(uber_data)
```
## Randomization Check
```{r}
summary(glm(treat ~ commute + period_month + period_weekday + period_hour, uber_data, family ="binomial"))
```

## Linear Regression
Since dependent variables are all categorical, we don't need to perform standardization.
### Model 1: target variable -> total trips
```{r}
uber_lm.total_trips <- lm(total_trips ~ treat + commute + period_month + period_weekday + period_hour, uber_data)
uber_lm.total_trips_int <- lm(total_trips ~ treat + commute + treat*commute + period_month + period_weekday + period_hour, uber_data)
uber_lm.total_trips_nohour <- lm(total_trips ~ treat + commute + period_month + period_weekday, uber_data)
uber_lm.total_trips_nohour_int <- lm(total_trips ~ treat + commute + treat*commute + period_month + period_weekday, uber_data)
uber_lm.total_trips_nocommute <- lm(total_trips ~ treat + period_month + period_weekday + period_hour, uber_data)
uber_lm.total_trips_commute_hour <- lm(total_trips ~ treat + commute+ period_month + period_weekday + period_hour + commute*period_hour, uber_data)
```

```{r}
library(stargazer)
stargazer(uber_lm.total_trips,uber_lm.total_trips_int,uber_lm.total_trips_nohour,uber_lm.total_trips_nohour_int,uber_lm.total_trips_nocommute,uber_lm.total_trips_commute_hour, type = 'html')
```
Model 6 has the highest adjusted R-squre. Apply AIC on two models that have the highest adj R-Square.
```{r}
library(car)
AIC(uber_lm.total_trips_int,uber_lm.total_trips_commute_hour)
```
```{r}
stargazer(uber_lm.total_trips_commute_hour, type="text")
```

The model with commute and treatment interaction term had a slightly higher AIC. Note that both models have very similar AIC scores.

With the models we have fitted previously, we notice:
- On average, with 5 min wait time, total trips tend to decrease
- However, wait time did not display high importance according to linear regression
- Whether trips happened at rush hours or not displayed high importance
- When the trips happened in a specific hour frame in a day seem to have high importance (daily seasonality in traffic demand)

Extension: Variable importance via ensemble methods
```{r}
library(randomForest)
rf_total_trips = randomForest(total_trips ~ treat + commute + period_month + period_weekday + period_hour, data=uber_data)
rf_total_trips$importance
```

### Model 2: target variable -> proportion of shared rides that were matched
```{r}
uber_lm.matched_prop <- lm(matched_prop ~ treat + commute + period_month + period_weekday + period_hour, uber_data)
uber_lm.matched_prop_int <- lm(matched_prop ~ treat + commute + treat*commute + period_month + period_weekday + period_hour, uber_data)
uber_lm.matched_prop_nohour <- lm(matched_prop ~ treat + commute + period_month + period_weekday, uber_data)
uber_lm.matched_prop_nohour_int <- lm(matched_prop ~ treat + commute + treat*commute + period_month + period_weekday, uber_data)
uber_lm.matched_prop_nocommute <- lm(matched_prop ~ treat + period_month + period_weekday + period_hour, uber_data)
uber_lm.matched_prop_commute_hour <- lm(matched_prop ~ treat + commute+ period_month + period_weekday + period_hour + commute*period_hour, uber_data)
```

```{r}
stargazer(uber_lm.matched_prop,uber_lm.matched_prop_int,uber_lm.matched_prop_nohour,uber_lm.matched_prop_nohour_int,uber_lm.matched_prop_nocommute,uber_lm.matched_prop_commute_hour, type = 'html')
```
Model 1 and model 2 had the highest R-square. Compare AIC:
```{r}
AIC(uber_lm.matched_prop,uber_lm.matched_prop_int)
```
Note that all models here didn't fit well. Adjusted R-square are below 0.4.

With the models we have fitted previously, we notice:
- With proportion of matched trips, wait time displayed high importance.
- With longer wait time, matched proportion would decrease
- Rush hours still displayed high importance
- Weekly and hourly seasonality displayed some importance
- If trips happened in March, matched proportion decreased compared to February

Extension: RF feature importance
```{r}
rf_matched_prop = randomForest(matched_prop ~ treat + commute + period_month + period_weekday + period_hour, data=uber_data)
rf_matched_prop$importance
```
### Model 3: target variable -> driver payout per trip
```{r}
uber_lm.avg_ride_payout <- lm(avg_ride_payout ~ treat + commute + period_month + period_weekday + period_hour, uber_data)
uber_lm.avg_ride_payout_int <- lm(avg_ride_payout ~ treat + commute + treat*commute + period_month + period_weekday + period_hour, uber_data)
uber_lm.avg_ride_payout_nohour <- lm(avg_ride_payout ~ treat + commute + period_month + period_weekday, uber_data)
uber_lm.avg_ride_payout_nohour_int <- lm(avg_ride_payout ~ treat + commute + treat*commute + period_month + period_weekday, uber_data)
uber_lm.avg_ride_payout_nocommute <- lm(avg_ride_payout ~ treat + period_month + period_weekday + period_hour, uber_data)
uber_lm.avg_ride_payout_commute_hour <- lm(avg_ride_payout ~ treat + commute+ period_month + period_weekday + period_hour + commute*period_hour, uber_data)
```

```{r}
stargazer(uber_lm.avg_ride_payout,uber_lm.avg_ride_payout_int,uber_lm.avg_ride_payout_nohour,uber_lm.avg_ride_payout_nohour_int,uber_lm.avg_ride_payout_nocommute,uber_lm.avg_ride_payout_commute_hour, type = 'html')
```
Note that models had low adj R-square (below 0.2). 

AIC on top 2 models:
```{r}
AIC(uber_lm.avg_ride_payout,uber_lm.avg_ride_payout_commute_hour)
```

With the models we have fitted previously, we notice:
- Avg driver payout demonstrated the efficiency of each trip
- More wait time on average decreased driver payout per trip.
- Rush hours and hourly seasonality displayed importance

Extension: Random Forest Feature Importance
```{r}
rf_avg_ride_payout = randomForest(avg_ride_payout ~ treat + commute + period_month + period_weekday + period_hour, data=uber_data)
rf_avg_ride_payout$importance
```
Note random forest emphasizes more on hourly seasonality.

## Decision Trees
Note that decision trees naturally could handle possible interactions between variable.
Note that R-square does not apply well with decision trees
### Model 1: target variable -> total trips
```{r}

dt_total_trips <- rpart(total_trips ~ treat + commute + period_month + period_weekday + period_hour, 
                  method = "anova", data = uber_data)
```

```{r}
dt_total_trips$variable.importance
```

Decision tree displayed no importance om treatment wrt total trips.

```{r}
print(dt_total_trips)
```

```{r}
plot(dt_total_trips, uniform = TRUE,
          main = "Total Trips with Decision Tree")
text(dt_total_trips, use.n = TRUE, cex = .7)
```
```{r}
plotcp(dt_total_trips)
```
Since our dependent variables are all categorical and we did not have a lot of nodes, there's no need to prune the tree.

```{r}
rmse_reg <- function(model_obj, testing = NULL, target = NULL) {
  #Calculates rmse for a regression decision tree
    yhat <- predict(model_obj, newdata = testing)
  actual <- testing[[target]]
  sqrt(mean((yhat-actual)^2))
}
rmse_reg(dt_total_trips,uber_data, "total_trips")
```

Compare to Linear regression
```{r}
library(caret)
linear_total_trip.10fold <- train(total_trips ~ treat + commute + treat*commute + period_month + period_weekday + period_hour, data = uber_data, trControl=trainControl(method="cv", number=10), method="lm")
print(linear_total_trip.10fold)
```
Decision tree has a higher lower RMSE. Due to the nature of the data set, it makes sense that decision trees performed better.

### Model 2: target variable -> proportion of shared rides that were matched
```{r}
dt_matched_prop <- rpart(matched_prop ~ treat + commute + period_month + period_weekday + period_hour, 
                  method = "anova", data = uber_data)
```

```{r}
dt_matched_prop$variable.importance
```

With matching proportion, treatment displayed some importance, not as much compared to other variables.

```{r}
print(dt_matched_prop)
```
This showed a few things that we didn't notice with linear regression
- If it is not rush hours, if wait time is 5 min, the matching rate is lower than that of 2 min
- If it is rush hours, demand is high and wait time does not matter
- Potential consideration: cancellation vs wait.

```{r}
plot(dt_matched_prop, uniform = TRUE,
          main = "Matching Proportion with Decision Tree")
text(dt_matched_prop, use.n = TRUE, cex = .7)
```
MSE comparison with linear regression
```{r}
rmse_reg(dt_matched_prop,uber_data, "matched_prop")
```

```{r}
linear_matched_prop.10fold <- train(matched_prop ~ treat + commute + treat*commute + period_month + period_weekday + period_hour, data = uber_data, trControl=trainControl(method="cv", number=10), method="lm")
print(linear_matched_prop.10fold)
```
Decision performed better.

### Model 3: target variable -> driver payout per trip
```{r}
dt_avg_ride_payout <- rpart(avg_ride_payout ~ treat + commute + period_month + period_weekday + period_hour, 
                            method = "anova", data = uber_data)
```

```{r}
dt_avg_ride_payout$variable.importance
```

Treatment displayed importance with decision tree, similar to linear regression result.
```{r}
print(dt_avg_ride_payout)
```
```{r}
plot(dt_avg_ride_payout, uniform = TRUE,
          main = "Driver Payout per Trip with Decision Tree")
text(dt_avg_ride_payout, use.n = TRUE, cex = .7)
```
- Similar to matching proportion, treatment doesn't affect driver payout per trip too much if it's rush hour.
- Rush hours on Tuesday, Thursday and Friday tends to have lower payout compare to rush hours on other days.
- If it's not rush hour, payout for longer wait time is lower.
- Wait time has some relationship with pricing.

Comparison with linear regression performance
```{r}
rmse_reg(dt_avg_ride_payout,uber_data, "avg_ride_payout")
```

```{r}
linear_avg_ride_payout.10fold <- train(avg_ride_payout ~ treat + commute + period_month + period_weekday + period_hour, data = uber_data, trControl=trainControl(method="cv", number=10), method="lm")
print(linear_avg_ride_payout.10fold)
```
### Check point: modified data
```{r}
write.csv(uber_data,'modified_uber_data.csv')
```

### Extension: Model 4 cancellation vs wait time
```{r}
dt_cancellations <- rpart(rider_cancellations/total_trips ~ treat + commute + period_month + period_weekday + period_hour, 
                          method = "anova", data = uber_data)
print(dt_cancellations)
```
```{r}
dt_cancellations$variable.importance
```

```{r}
uber_data['cancellation_rate'] = uber_data$rider_cancellations/uber_data$total_trips
treatment_uber['cancellation_rate'] = treatment_uber$rider_cancellations/treatment_uber$total_trips
control_uber['cancellation_rate'] = control_uber$rider_cancellations/control_uber$total_trips
```


```{r}
ggplot(treatment_uber,aes(period_start,cancellation_rate))+geom_line(aes(color="cancellation_rate_5min"))+geom_line(data=control_uber,aes(color="cancellation_rate_2min"))+labs(color="service type")+geom_point(data=treatment_uber[treatment_uber$commute==TRUE,],aes(period_start,cancellation_rate))+ geom_point(data=control_uber[control_uber$commute==TRUE,],aes(period_start,cancellation_rate))
```


### Extension: Model 5 pool trips vs wait time
```{r}
dt_pool <- rpart(trips_pool ~ treat + commute + period_month + period_weekday + period_hour, 
                          method = "anova", data = uber_data)
print(dt_pool)
print(dt_pool$variable.importance)
```
```{r}
plot(dt_pool, uniform = TRUE,
          main = "Pool Trips with Decision Tree")
text(dt_pool, use.n = TRUE, cex = .7)
```


### Extension: Model 6 express trips vs wait time
```{r}
dt_trips_express <- rpart(trips_express ~ treat + commute + period_month + period_weekday + period_hour, 
                          method = "anova", data = uber_data)
print(dt_trips_express)
print(dt_trips_express$variable.importance)
```
```{r}
plot(dt_trips_express, uniform = TRUE,
          main = "Express Trips with Decision Tree")
text(dt_trips_express, use.n = TRUE, cex = .7)
```

